---
title: "Bioconductor_GenomicDataScience_Quiz1"
author: "Ojaswini Pandey"
date: "2023-04-15"
output: html_document
---

```{r}

#loading all the libraries

library(IRanges)
library(GenomicRanges)
library(dbplyr)
#library(BiofileCache)
library(rtracklayer)
library(AnnotationHub)
library(GenomeInfoDb)


```

 
Q1 : Use the AnnotationHub package to obtain data on "CpG Islands" in the human genome. How many islands exists on the autosomes?

```{r}

#getting the record for human cpg island
ahub = AnnotationHub()
ahub_human = subset(ahub, species == "Homo sapiens")  
ahub_human_cpg = query(ahub_human, "CpG Islands")
ahub_human_cpg
ahub_human_cpg_record = ahub_human_cpg[["AH5086"]]
ahub_human_cpg_record

#creating a filter to exract autosomes
filter = c(paste("chr", 1:22, sep =""))

#spliting the record obtatained for human cpg islands
split_ahub_human_cpg_record = split(ahub_human_cpg_record, seqnames(ahub_human_cpg_record))

#applying filter on the split record
autosomes = split_ahub_human_cpg_record[filter]

#finding the number of autosomes
unlist(autosomes)

```

Q2 : How many CpG Islands exists on chromosome 4?

```{r}
autosomes[4]
```


Q3 : Obtain the data for the H3K4me3 histone modification for the H1 cell line from Epigenomics Roadmap, using AnnotationHub. Subset these regions to only keep regions mapped to the autosomes (chromosomes 1 to 22).

```{r}
# Obtaining the data for H3K4me3 histone modification for the H1 cell line from the Epigenomics Roadmap using AnnotationHub.
ahub_H3K4me3 = query(ahub, c("H3K4me3", "narrowpeak", "E003"))
ahub_H3K4me3

# Retrieving the specific record
ahub_H3K4me3_record = ahub_H3K4me3[["AH29884"]]
ahub_H3K4me3_record

# Extracting Autosomes by applying the filter created earlier
split_ahub_H3K4me3_record = split(ahub_H3K4me3_record, seqnames(ahub_H3K4me3_record))
autosomes_ahub_H3K4me3 = split_ahub_H3K4me3_record[filter]
autosomes_ahub_H3K4me3

#Finding the number of bases these regions cover
sum(width(unlist(autosomes_ahub_H3K4me3)))

```
Q4 : Obtain the data for the H3K27me3 histone modification for the H1 cell line from Epigenomics Roadmap, using the AnnotationHub package. Subset these regions to only keep regions mapped to the autosomes. In the return data, each region has an associated “signalValue”. What is the mean signalValue across all regions on the standard chromosomes?

```{r}
# Obtaining the data for H3K27me3 histone modification for the H1 cell line from the Epigenomics Roadmap using AnnotationHub.
ahub_H3K27me3 = query(ahub, c("H3K27me3", "narrowpeak", "E003"))
ahub_H3K27me3

# Retrieving the specific record
ahub_H3K27me3_record = ahub_H3K27me3[["AH29892"]]
ahub_H3K27me3_record

# Extracting Autosomes by applying the filter created earlier
split_ahub_H3K27me3_record = split(ahub_H3K27me3_record, seqnames(ahub_H3K27me3_record))
autosomes_ahub_H3K27me3 = split_ahub_H3K27me3_record[filter]
autosomes_ahub_H3K27me3

# Creating a subset of autosomes
autosomes1_ahub_H3K27me3 = subset(ahub_H3K27me3_record, seqnames %in% filter)

# finding mean signal value
mean_signalValue = mean(autosomes1_ahub_H3K27me3$signalValue)
mean_signalValue

```

Q5 : Bivalent regions are bound by both H3K4me3 and H3K27me3. Using the regions we have obtained above, how many bases on the standard chromosomes are bivalently marked?

```{r}

# Finding the intersection between the two histone modifications for autosomes
biValent = intersect(unlist(autosomes_ahub_H3K4me3), unlist(autosomes_ahub_H3K27me3))
sum(width(biValent))

```

Q6 : We will examine the extent to which bivalent regions overlap CpG Islands. How big a fraction (expressed as a number between 0 and 1) of the bivalent regions, overlap one or more CpG Islands?
 
```{r}
# finding overlap of bivalent regions with CpG islands
cpg_bivalent = findOverlaps(biValent, unlist(autosomes))
cpg_bivalent

# calculating the fraction of the bivalent regions overlap with CpG islands
fraction_biValent = length(unique(queryHits(cpg_bivalent)))/length(biValent)
fraction_biValent

```
Q7 : How big a fraction (expressed as a number between 0 and 1) of the bases which are part of CpG Islands, are also bivalent marked?

```{r}
cpg_bivalent_intersect = intersect(biValent, unlist(autosomes))

#calculating the fraction of bivalent bases which are part of CpG islands
fraction_cpg_bivalent_intersect = sum(width(reduce(cpg_bivalent_intersect)))/sum(width(unlist(autosomes)))
fraction_cpg_bivalent_intersect

```

Q8 : How many bases are bivalently marked within 10kb of CpG Islands? Tip: consider using the “resize()”" function.

```{r}
cpg_10kb <- resize(unlist(autosomes), width = 20000 + width(unlist(autosomes)), fix = "center")
cpg_10kb_bivalent <- intersect(cpg_10kb, biValent)
sum(width(cpg_10kb_bivalent))


```


Q9 : 9. How big a fraction (expressed as a number between 0 and 1) of the human genome is contained in a CpG Island?
Tip 1: the object returned by AnnotationHub contains “seqlengths”.
Tip 2: you may encounter an integer overflow. As described in the session on R Basic Types, you can address this by converting integers to numeric before summing them, “as.numeric()”.


```{r}

# calculating human genome size

chr_list = c(paste("chr", 1:22, sep=""))
genome = keepSeqlevels(ahub_human_cpg_record, chr_list, pruning.mode = "coarse")
genome_size = sum(as.numeric(seqlengths(genome)))

# calculating the fraction of human genome which contained CpG Island
cpg_autosomes_size = sum(as.numeric(width(unlist(autosomes))))
cpg_autosomes_size / genome_size

```


Q10 : Compute an odds-ratio for the overlap of bivalent marks with CpG islands.

```{r}

# calculating InOut matrix
overlapMat = matrix(0, ncol = 2, nrow = 2)
colnames(overlapMat) <- c("in", "out")
rownames(overlapMat) <- c("in", "out")
overlapMat[1,1] <- sum(width(cpg_bivalent_intersect))
overlapMat[1,2] <- sum(width(setdiff(biValent, unlist(autosomes))))
overlapMat[2,1] <- sum(width(setdiff(unlist(autosomes), biValent)))
overlapMat[2,2] <- genome_size - sum(overlapMat)

# calculating odds-ratio
oddsRatio <- overlapMat[1,1] * overlapMat[2,2] / (overlapMat[2,1] * overlapMat[1,2])
oddsRatio


```
```{r}
sessionInfo()
```


